---
title: "Maryland IBV Lineage Phylogenetics Comparison to 2022"
description: "GISAID Collection of IBV strains from B/Victoria and B/Yamagata" 
output: html_notebook
---

# IBV Lineage Comparison

## Data collection

Sequence were aquired from GISAID in August of 2022 with the following filters: 
  - Segments: HA, NS
  - Dates: All until 2022
  - Complete Sequences only
  
## Sequence alignment and tree building

Sequences were aligned using ClustalO. Phylogenetic trees were constructed using RAxML 

## Plot using ggtree

```{r, install ggtree}

#ggtree Test 
#install 
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("YuLab-SMU/treedataverse") #includes ggtree along with other essential packages

# browseVignettes("ggtree") #essnetial guide for tree browsing
```

https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html
https://yulab-smu.top/treedata-book/chapter13.html
https://yulab-smu.top/treedata-book/faq.html#faq-local-file
https://yulab-smu.top/treedata-book/chapter12.html
https://www.biostars.org/p/312389/
https://groups.google.com/g/bioc-ggtree/c/Ciy4Fh_7hl4
https://bioconductor.statistik.tu-dortmund.de/packages/3.6/bioc/vignettes/ggtree/inst/doc/treeAnnotation.html
https://bioconnector.github.io/workshops/r-ggtree.html#labeling_clades

# Building HA Tree
B/Victoria vs B/Yamagata

Still missing B/Yamagata Sequences
  - R0337 
  - R0225

Get from Jo

```{r, IBVHA nucleotide tree - trial}
library(knitr)
library(treedataverse)
library(ggtree)
library(ggimage)
library(tidytree)

#Importing newick file
HAnewick <- read.newick("B_Vic_B_Yam_HA_Maryland_epiflu_isolates_gisaid_NJ_cs.newick", node.label = "label") #import newick tree
print(HAnewick)

pHA <-  ggtree(HAnewick)

pHA_annot <- pHA + 
  geom_cladelab(node=253, label="B/Victoria", color='red', align = TRUE, barsize = 1, align=TRUE, angle=270, hjust='center', offset.text=.002) +
  theme_tree2() + 
  geom_cladelab(node=358, label="B/Yamagata", color='blue', barsize = 1, align=TRUE, angle=270, hjust='center', offset.text= 0.002, offset= 0, fontsize = 4 )


pHA_annot 


#geom_text(aes(label=node)) #adds node number to identify

#nodes
#251 - Vic vs Yam
# 393 - yam
# 252 - vic


```


```{r, IBV HA NJ tree with 4/4 Study Strains}

library(knitr)
library(treedataverse)
library(ggtree)
library(ggimage)
library(tidytree)
library(ggrepel)

HA_NJ_newick <- read.newick("B_Vic_B_Yam_HA_Maryland_epiflu_isolates_gisaid_NJ_cs.newick", node.label = "label") #import newick tree of RAXML HA build
print(HA_NJ_newick)

pHA_NJ <-  ggtree(HA_NJ_newick)

pHA_lim <- pHA_NJ + 
  geom_treescale() +
  #Yamagata
  geom_cladelab(node=288,
                label="Yamagata",
                barcolour='red',
                offset = -0.08,
                offset.text = 0.003,
                barsize = 1, 
                align=TRUE, 
                angle=270, 
                hjust='center',
                fontsize = 6) +
  #Victoria  
  geom_cladelab(node=375, 
                label="Victoria", 
                barcolour='blue', 
                barsize = 1,
                offset = 0.01495,
                offset.text = 0.003,
                align=TRUE, 
                angle=270, 
                hjust='center', 
                fontsize = 6) + 
  geom_tiplab(aes(subset=(grepl("'B/Baltimore/R0122/2016'",label,fixed=TRUE)==TRUE)), align = TRUE, offset=0.001, size = 1.5) + 
  geom_tiplab(aes(subset=(grepl("'B/Baltimore/R0001/2016'",label,fixed=TRUE)==TRUE)), align = TRUE, offset=0.001, size = 1.5 ) + 
  geom_tiplab(aes(subset=(grepl("'B/Baltimore/R0250/2016'",label,fixed=TRUE)==TRUE)), align = TRUE, offset=0.001, size = 1.5 ) + 
  theme_tree() 
  # geom_text(aes(label=node), size = 1.6 ) #labels branch node ID for labeling purposes...bc Im lazy


pHA_lim #plot tree

#ggsave("B_Maryland_HA_tree_FINAL.jpeg", width = 7, height = 7) #save plot to relative dir


# B/Yam - node 252
# B/Vic - node 353
```



# Building NS Tree
B/Victoria vs B/Yamagata
RAXML tree chosen for cluster Clarity



```{r IBV NS nucleotide tree}

library(knitr)
library(treedataverse)
library(ggtree)
library(ggimage)
library(tidytree)

#Importing newick file
NSnewick <- read.newick("B_Vic_B_Yam_NS_Maryland_epiflu_isolates_gisaid_RAxML.newick", node.label = "label") #import newick tree

# node 350
# node 250

print(NSnewick)

pNS_RAXML <-  ggtree(NSnewick)

pNS_lim <- pNS_RAXML + 
  geom_treescale() +
  #Yamagata
  geom_cladelab(node=250,
                label="Yamagata",
                barcolour='red',
                offset = 0,
                offset.text = 0.003,
                barsize = 1, 
                align=TRUE, 
                angle=270, 
                hjust='center',
                fontsize = 6) +
  #Victoria 
  geom_cladelab(node=350, 
                label="Victoria", 
                barcolour='blue', 
                barsize = 1,
                offset = 0.01495,
                offset.text = 0.003,
                align=TRUE, 
                angle=270, 
                hjust='center', 
                fontsize = 6)

#'''#for labeling specific samples when sequences are avaialble
#  geom_tiplab(aes(subset=(grepl("B/Baltimore/R0122/2016",label,fixed=TRUE)==TRUE)), align = TRUE, offset=0.001, size = 1.5) + #annotated strains
#  geom_tiplab(aes(subset=(grepl("B/Baltimore/R0001/2016",label,fixed=TRUE)==TRUE)), align = TRUE, offset=0.001, size = 1.5 ) + #annotated strains 
#  theme_tree() 


pNS_lim #plot tree

ggsave("B_Maryland_NS_tree.jpeg", width = 7, height = 7) #save plot to relative dir



```

```{r IBV RAXML NS nucleotide tree with 3/4 IBV clinical isolates}
library(knitr)
library(treedataverse)
library(ggtree)
library(ggimage)
library(tidytree)

#Importing newick file
NSnewick_all <- read.newick("B_Vic_B_Yam_NS_Maryland_epiflu_isolates_gisaid_RAxML_cs.newick", node.label = "label") #import newick tree

# node 350
# node 250

print(NSnewick_all)

pNS_all_RAXML <-  ggtree(NSnewick_all)

pNS_all_lim <- pNS_all_RAXML + 
  geom_treescale() +
  #Yamagata
  geom_cladelab(node=394,
                label="Yamagata",
                barcolour='red',
                offset = 0,
                offset.text = 0.003,
                barsize = 1, 
                align=TRUE, 
                angle=270, 
                hjust='center',
                fontsize = 6) +
  #Victoria 
  geom_cladelab(node=253, 
                label="Victoria", 
                barcolour='blue', 
                barsize = 1,
                offset = 0,
                offset.text = 0.003,
                align=TRUE, 
                angle=270, 
                hjust='center', 
                fontsize = 6) + 
  geom_tiplab(aes(subset=(grepl("B/Baltimore/R001/2016",label,fixed=TRUE)==TRUE)), align = TRUE, offset=-0.02, size = 3) + 
  geom_tiplab(aes(subset=(grepl("B/Baltimore/R0122/2016",label,fixed=TRUE)==TRUE)), align = TRUE, offset=-0.02, size = 3) +
  geom_tiplab(aes(subset=(grepl("B/Baltimore/R0250/2016",label,fixed=TRUE)==TRUE)), align = TRUE, offset=-0.02, size = 3) + 
  theme_tree() 
  #geom_text(aes(label=node)) #adds node number to identify


pNS_all_lim #plot tree

ggsave("B_Maryland_NS_tree_ALL_FINAL.jpeg", width = 7, height = 7) #save plot to relative dir

# Node 394 - Yam 
# Node 253 - VIc
```








